Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Updates to paper, now includes figures

\begin{enumerate}
  \item Identify the domain of interest. (Section~\ref{SecIdentifyDomain})
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \cite{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Interview the developers. (Section~\ref{SecSurvey})
	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
	\item Analyze the results and answer the research questions. (Sections~\ref{repmetrics}---\ref{Sec_Recommendations}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
\end{enumerate}

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interview developers
using a list of 20 questions \cite{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask them about
their understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
are about specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary.  Each interview should take about 1 hour.

The interviewees should follow a process where they can make informed consent.
The interviews should follow standard ethics guideline of asking for consent
before interviewing, recording, and including participant details in the report.
The interview process presented here was approved by the McMaster University
Research Ethics Board under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.  For LBM we were able to recruit 4 developers to participate in
our study. Results in next sections.

@@

Since each domain we will study will have a reasonably small scope, we will be
able to view the software as constituting a program family. Studying the common
properties within a family of related programs is termed a domain analysis.  For
the current methodology time constraints necessitate a shallow domain analysis.
A table should be constructed that distinguishes the programs under study by the
variabilities that distinguish them.  In research software the variabilities are
often related to assumptions.  Table~\ref{tbl_features} shows the variabilities
for the LBM software example.

we will
contrast the different packages based on their features, as shown in
Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
software product line.  The packages all have the commonality of using LBM
techniques to simulate the motion of fluids, but there are also variabilities
between them.  LBM packages can be distinguished by the dimension of problems
they simulate, whether or not (and how) they support parallel computation,
whether or not compressibility can be considered, whether reflexive boundaries
can be employed, whether the simulation handles multi-fluids, whether the
simulations handles turbulence, and whether complex geometries can be simulated.
All of these variabilities are listed as columns in Table~\ref{tbl_features}
with the software listed in alphabetical order. A more complete analysis of the
commonalities and variabilities between LBM packages is provided in
\cite{Michalski2021}.

\begin{table}[h!]
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\toprule
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\midrule
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\bottomrule
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
			Complex Geometries (Yes or No), OS for Operating System (Windows
			(W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
practices to the communities ranking.  Our best practices ranking comes from the
results of the AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the
communities ranking by repository stars and watches.  The comparison will
provide insight on whether best practices are rewarded by popularity.  However,
inconsistencies between the AHP ranking and the communities ranking are in
inevitable for the following reasons: 
\begin{inparaenum}[i)]
	\item the overall quality ranking via AHP makes the unrealistic assumption
	of equal weighting between the different quality factors;
	\item stars are known to not be a particularly good measure of popularity,
	since young projects have less time to accumulate stars \cite{}; 
	\item and, as for consumer products, there are more factors that determine
	popularity than quality alone.
\end{inparaenum}

Table~\ref{repometrics} compares the AHP ranking of the LBM package versus their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a measure of repository stars. Looking at the repository stars of
the other 15 packages, we can observe a pattern where packages that have been
highly ranked by our assessment tend to have more stars than lower ranked
packages. The best ranked package by AHP (ESPResSo) has the second most stars,
while the ninth ranked package (Sailfish) has the highest number of stars. The
same correlation is observed in the repository watch column, although this
column contains less data, since two of the packages (Palabos, waLBerla) use
GitLab, which does not track watches. Packages designated as lower quality often
do not use GitHub or GitLab, or have only a few stars/watches. Although the AHP
ranking and the community popularity estimate are not perfect measures, they do
suggest a correlation between best practices and popularity. We do not know
which comes first, the use of best practices or popularity, but we do know that
the top ranked packages tend to incorporate best practices.

@@

In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
comparing the artifacts that we observed in LBM repositories to those observed
and recommended for research software in general.  As part of filling in the
measurement template (Section~\ref{}), the domain software is examined for the
presence of artifacts, which are then categorized by frequency. We suggest
grouping them into the categories: common, less common, and rare.  Common
artifacts are those found in more than two thirds of the examples, less common
artifacts are between 1/3 and 2/3, while rare are in less than 1/3 of the
software packages.

The observed frequency of artifacts should then be compared to the artifacts
recommended by research software guidelines.  Table~\ref{} summarizes the
artifacts recommended by some existing guidelines.

@@

The results for the LBM example are shown below.  The majority of LBM generated
artifacts correspond to general recommendations from research software
developers.  For LBM, a union of the three categories mostly corresponds to
recommendations made by the research software community, as shown in
Table~\ref{Tbl_Guidelines}.

@@

Although the LBM community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed as follows: API documentation, a roadmap, a code of
conduct, programming style guide, uninstall instructions.

@@

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts \cite[p.\
501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.  In this section we answer
\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
to their utilization in the research software community in general.

\begin{table}[h!]
\begin{center}
\begin{tabular}{ p{12 cm}}
%\toprule
\textbf{Development Tools}\\
\midrule
Continuous Integration, Code Editors, Development Environments, Runtime
Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
%\midrule
\textbf{Dependencies}\\
\midrule
Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
%\midrule
\textbf{Project Management Tools}\\
\midrule
Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
%\bottomrule
\end{tabular}
\caption{Observed development tools, dependencies and project management tools} 
\label{tbl_tools}
\end{center}
\end{table}

@@

Based on interviews with 4 developers, this section aims to answer the research
questions: i) What are the pain points for developers working on research
software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
developers from LBM compare to the pain points for research software in general
(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
points and include citations that contrast the LBM experience with observations
from researchers in other domains.  Potential ways to address the pain points
are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
are found in \cite{SmithEtAl2021}.

>>>>

\begin{enumerate}
  \item Identify the domain of interest. (Section~\ref{SecIdentifyDomain})
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \cite{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
	\item Interview the developers. (Section~\ref{SecSurvey})
    \item Domain analysis.  (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions. (Sections~\ref{repmetrics}---\ref{painpoints}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
\end{enumerate}

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies} and \rqref{RQ_PainPoints}) require going beyond
the quantitative data from the measurement template. To gain the required
insight, we interview developers using a list of 20 questions
\cite{SmithEtAl2021}. The questions cover the background of the development
teams, the interviewees, and the software itself. We ask the developers how they
organize their projects. We also ask them about their understanding of the
users. Some questions focus on the current and past difficulties, and the
solutions the team has found, or will try. We also discuss the importance of,
and the current situation for, documentation. A few questions are about specific
software qualities, such as maintainability, understandability, usability, and
reproducibility. The interviews are semi-structured based on the question list;
we ask follow-up questions when necessary.  Each interview should take about 1
hour.

The interviewees should follow a process where they can make informed consent.
The interviews should follow standard ethics guideline of asking for consent
before interviewing, recording, and including participant details in the report.
The interview process presented here was approved by the McMaster University
Research Ethics Board under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.  For LBM we were able to recruit 4 developers to participate in
our study.

@@

Since each domain we will study will have a reasonably small scope, we will be
able to view the software as constituting a program family. Studying the common
properties within a family of related programs is termed a domain analysis.  For
the current methodology time constraints necessitate a shallow analysis. For
each domain a table should be constructed that distinguishes the programs under
study by their variabilities.  In research software the variabilities are often
related to assumptions.  Table~\ref{tbl_features} shows the variabilities for
the LBM software example~\cite{Michalski2021}.

\begin{table}[h!]
	\begin{center}
		\begin{tabular}{ p{3cm}p{1.25cm}p{2.25cm}llllll}
			\toprule
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\midrule
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\bottomrule
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
			Complex Geometries (Yes or No), OS for Operating System (Windows
			(W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
practices to the communities ranking.  Our best practices ranking comes from the
results of the AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the
communities ranking by repository stars and watches.  The comparison will
provide insight on whether best practices are rewarded by popularity.  However,
inconsistencies between the AHP ranking and the communities ranking are in
inevitable for the following reasons: 
\begin{inparaenum}[i)]
	\item the overall quality ranking via AHP makes the unrealistic assumption
	of equal weighting between the different quality factors;
	\item stars are known to not be a particularly good measure of popularity,
	because of how people use stars and since young projects have less time to
	accumulate stars~\cite{Szulik2017}; 
	\item and, as for consumer products, there are more factors that determine
	popularity than quality alone.
\end{inparaenum}

Table~\ref{repometrics} compares the AHP ranking of the LBM package versus their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a measure of repository stars. Looking at the repository stars of
the other 15 packages, we can observe a pattern where packages that have been
highly ranked by our assessment tend to have more stars than lower ranked
packages. The best ranked package by AHP (ESPResSo) has the second most stars,
while the ninth ranked package (Sailfish) has the highest number of stars. The
same correlation is observed in the repository watch column, although this
column contains less data, since two of the packages (Palabos, waLBerla) use
GitLab, which does not track watches. Although the AHP ranking and the community
popularity estimate are not perfect measures, they do suggest a correlation
between best practices and popularity.

@@

In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
comparing the artifacts that we observed in the domain repositories to those
observed and recommended for research software in general.  While filling in the
measurement template (Section~\ref{empiricalmeasures}), the domain software is
examined for the presence of artifacts, which are then categorized by frequency.
We suggest grouping them into the categories: common (more than 2/3 of
projects), less common (between 1/3 and 2/3), and rare (less than 1/3 of
projects). The observed frequency of artifacts should then be compared to the
artifacts recommended by research software guidelines, as summarized in
Table~\ref{Tbl_Guidelines}.

@@

The results for the LBM example are shown below.  The majority of LBM generated
artifacts correspond to general recommendations from research software
developers.  For LBM, a union of the three categories mostly corresponds to
recommendations made by the research software community, as shown in
Table~\ref{Tbl_Guidelines}. Although the LBM community participates in most of
the practices we found listed in the general research software guidelines, some
recommended practices were not observed or rarely observed as follows: API
documentation, a roadmap, a code of conduct, programming style guide, uninstall
instructions.

@@

@@

To answer \rqref{RQ_CompareToolsProjMngmnt} we need to summarize tools that are
visible in the repositories and that are mentioned during the developer
interviews. Software tools are used to support the development, verification,
maintenance, and evolution of software, software processes, and
artifacts~\cite[p.\ 501]{GhezziEtAl2003}. Development tools support the
development of end products, but do not become part of them, unlike dependencies
that remain in the application once it is released \cite[p.\
506]{GhezziEtAl2003}. The tools found for LBM software packages are as follows:

\begin{description}
	\item[Development Tools:] Continuous Integration, Code Editors, Development
	Environments, Runtime Environments, Compilers, Unit Testing Tools,
	Correctness Verification Tools
	\item[Dependencies:] Build Automation Tools, Technical Libraries, Domain
	Specific Libraries
	\item[Project Management Tools:] Collaboration Tools, Email, Change Tracking
	Tools, Version Control Tools, Document Generation Tools
\end{description}

% \begin{table}[h!]
% \begin{center}
% \begin{tabular}{ p{12 cm}}
% %\toprule
% \textbf{Development Tools}\\
% \midrule
% Continuous Integration, Code Editors, Development Environments, Runtime
% Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
% %\midrule
% \textbf{Dependencies}\\
% \midrule
% Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
% %\midrule
% \textbf{Project Management Tools}\\
% \midrule
% Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
% %\bottomrule
% \end{tabular}
% \caption{Observed development tools, dependencies and project management tools} 
% \label{tbl_tools}
% \end{center}
% \end{table}

Once the data on tools is collected, two specific tools that can be compared to
research software norms are version control and continuous integration.

@@

Ask developers during the interviews about their pain points.  Compare the identified pain points to the literature that discusses pain points \cite{Wieseet}.
Based on interviews with 4 developers, this section aims to answer the research
questions: i) What are the pain points for developers working on research
software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
developers from LBM compare to the pain points for research software in general
(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
points and include citations that contrast the LBM experience with observations
from researchers in other domains.  Potential ways to address the pain points
are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
are found in \cite{SmithEtAl2021}.
